Testing Charts: viewers' perceptual accuracy in surveys

### 2023-07-24 SDSS-2023/testing-charts-accuracy.qmd
results for 3b

```{=tex}
\hh{Not sure where should put this, parking it here for right now:}
\hh{expectations - start}
```
What we call 'aligned' and 'unaligned', here, is similar to Cleveland and McGill's positions along aligned and unaligned scales, but with some modifications: both 'aligned' and 'unaligned' bars share the same axis.
Aligned tiles are additionally anchored in the same position in one dimension, i.e. the difference between their sizes can be reduced to a positional assessment, which in the absence of three-dimensional cues seems to be the general strategy for evaluating barcharts [@zacksReadingBarGraphs1998a].  
Unaligned tiles do not share this anchor, however, the context of the other tiles in the chart provide a frame, which *should* help with an assessment of the tiles' sizes.

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the context of the other tiles in the same column mitigating some of this difficulty.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.
\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}
\hh{What about the sketch now?}

@@

```{=tex}
\hh{I think we should include the floating bars here to demonstrate that they are worse than the wide horizontal bars. }
```

@@

\hh{expectations - end}

@@

<!--
```{r eval=FALSE}
#| label: fig-tasks-old
#| fig-cap: "Each of the visual stimuli presented to viewers. In each bar chart, two pieces are marked. The larger piece in each chart are pieces B and D. "
#| fig-subcap:
#| - "Vertical, unaligned"
#| - "Horizontal, unaligned"
#| - "Wide horizontal, unaligned"
#| - "Vertical, aligned"
#| - "Horizontal, aligned"
#| - "Wide horizontal, aligned"
#| layout-nrow: 3
#| out-width: 90%
cat("Task 1")
cat("Task 2")
cat("Task 3")
knitr::include_graphics("images/vertical-ab.PNG")
knitr::include_graphics("images/round3/horizontal-ab.PNG")
knitr::include_graphics("images/round3/hwide-ab.PNG")
knitr::include_graphics("images/vertical-cd.PNG")
knitr::include_graphics("images/round3/horizontal-cd.PNG")
knitr::include_graphics("images/round3/hwide-cd.PNG")
```
-->

@@

@@

| Name    | Description    |      \# Participants |                effective sample size |
|---------------|---------------|--------------:|--------------:|
| Task 1 | Vertical| `r nrow(round1_adj) + nrow(round2_adj)` | `r round(neff(round_12$weight),1)` |
| Task 2 | Horizontal |`r nrow(round3_adj[round3_adj$P_EXP == 1,])` | `r round(neff(round3_adj[round3_adj$P_EXP == 1,]$weight),1)` |
| Task 3 | Horizontal wide |`r nrow(round3_adj[round3_adj$P_EXP == 2,])` | `r round(neff(round3_adj[round3_adj$P_EXP == 2,]$weight),1)` |

@@

One pattern in accuracy holds across each of the three structural variations: the aligned task has a higher level of accuracy than its unaligned counterpart. Interestingly, while we expect an improvement in accuracy when shifting from the horizontal to the horizontal wide design given the larger difference in pixel length between the two pieces, the resulting effects on the accuracy of the responses are not completely straightforward: the shift from a vertical to the (tall) horizontal design is detrimental to an accurate perception for both aligned and unaligned comparisons. The re-scaled design of the wide horizontal bars reclaims some of the loss for unaligned bars and outperforms the vertical design by a similar margin in aligned bars, but does not out-perform the vertical design when comparing unaligned tiles.

@@

>>>>

```{=tex}
\hh{Not sure where should put this, parking it here for right now:}
\hh{expectations - start}
```
What we call 'aligned' and 'unaligned', here, is similar to Cleveland and McGill's positions along aligned and unaligned scales, but with some modifications.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.

In this paper, both 'aligned' and 'unaligned' bars share the same axis.

Aligned tiles are additionally anchored in the same position along the axis, i.e. the difference between their sizes can be reduced to a positional assessment, which in the absence of three-dimensional cues seems to be the general strategy viewers use when evaluating barcharts [@zacksReadingBarGraphs1998a].  
Unaligned tiles do not share this anchor of position, i.e. the difference between their sizes needs at least two comparisons of positions.
However, unaligned bars in a stacked barchart are not just assessments of the lengths of bars -- the other tiles in the chart provide a reference frame of the shared axis, which *should* help with an assessment of the tiles' sizes. As shown in the sketch of @fig-tasks-explain, this framing leads to a stimulus that is between Cleveland and McGill's task of 'position along the same but unaligned axis' and an assessment of length.

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the other tiles in the same column mitigating some of this difficulty. \hh{In particular, the additional context of the other tiles should *help* rather than *hurt* the comparison, unlike other situations, where additional information besides the pieces involved in the comparison were found to be distractors} [@talbotFourExperimentsPerception2014].
<!--\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}-->
<!--\hh{What about the sketch now?}-->

@@

@@

@@

@@

### Task 3b: Floating bars

```{r}
n5 <- nrow(round5_adj)
n5_eff <- neff(round5_adj$weight)
```

\hh{
In order to measure the impact of the chart's context, we asked a subset of participants to assess the size of the two tiles A and B (as shown in the bottom right of @fig-tasks-explain) to each other without giving any additional information.}

@@

| Name    | Description    |      \# Participants |                effective sample size |
|---------------|---------------|--------------:|--------------:|
| Task 1 | Vertical| `r nrow(round1_adj) + nrow(round2_adj)` | `r round(neff(round_12$weight),1)` |
| Task 2 | Horizontal |`r nrow(round3_adj[round3_adj$P_EXP == 1,])` | `r round(neff(round3_adj[round3_adj$P_EXP == 1,]$weight),1)` |
| Task 3 | Horizontal wide |`r nrow(round3_adj[round3_adj$P_EXP == 2,])` | `r round(neff(round3_adj[round3_adj$P_EXP == 2,]$weight),1)` |
| Task 3b | Floating bars |`r n5` | `r sprintf("%.1f",n5_eff)` |

@@

One pattern in accuracy holds across each of the three structural variations: the aligned task has a higher level of accuracy than its unaligned counterpart. Interestingly, while we expect an improvement in accuracy when shifting from the horizontal to the horizontal wide design given the larger difference in pixel length between the two pieces, the resulting effects on the accuracy of the responses are not completely straightforward: the shift from a vertical to the (tall) horizontal design is detrimental to an accurate perception for both aligned and unaligned comparisons. The re-scaled design of the wide horizontal bars reclaims some of the loss for unaligned bars and outperforms the vertical design by a similar margin in aligned bars, but does not out-perform the vertical design when comparing unaligned tiles. 
\hh{The beneficial effect of frames/context of the chart is apparent when comparing accuracy of responses for the Horizontal wide design and the floating bars: the rate of correct responses in floating bars is significantly lower than in the wide horizontal charts. }

@@

\hh{The actionable insights are:}

- make sure that the main comparison in a plot is lined up -- in an interactive setting it would be beneficial to have the possibility of re-ordering the categories to help with comparisons. 
- offering zoom-ins help with certainty, even if it does not help with perceptual accuracy
